\begin{abstract}
Protein--protein binding affinity prediction is an attractive target for machine learning because affinity is experimentally measurable, biologically important, and directly relevant to therapeutic design. At the same time, it is technically difficult: affinity depends on a three-dimensional interface, long-range structural context, solvent exposure, and conformational dynamics, while available structure--affinity datasets remain comparatively small. This thesis investigates whether short explicit-solvent molecular dynamics simulations can improve residue-graph affinity prediction when compared with static structural features alone.

The active pipeline begins with a curated wild-type subset of the PPB-Affinity dataset and processes each complex through explicit-solvent molecular dynamics, residue-level graph construction with DeepRank2, and downstream learning with either a masked graph variational autoencoder or a direct supervised graph regressor. Two graph views are considered: a full-complex residue graph and a fixed-size interface patch. Two feature modes are also compared: a static representation ($S$) and a static-plus-dynamic representation ($SD$), where the dynamic channels summarize hydration, C$\alpha$ flexibility, backbone torsions, contact persistence, distance statistics, and approximate inter-chain nonbonded force proxies. Final evaluation is performed on a matched 1437-complex subset using a repeated outer resampling design with 52 model configurations, 5 folds, 2 repeats, and 520 total fits.

The confirmatory results are clear. The full graph view outperforms the interface view in all 24 matched variational-model comparisons, with a mean interface penalty of 0.1005 test-RMSE units. The dynamic augmentation does not yield a robust held-out gain over the static representation; in 11 of 12 matched $S$ versus $SD$ comparisons under the shared-static reconstruction policy, the static model performs better. The strongest overall model is the direct supervised baseline on the full static graph, with mean test RMSE 2.3379, mean test Pearson correlation 0.5064, and mean test $R^2$ 0.2146. Semi-supervised dynamic models fit the training data more strongly but also show the largest train--test gaps, indicating increased flexibility without a commensurate gain in generalization.

The main conclusion is therefore not that protein dynamics are unimportant, but that the present dynamic encoding strategy is not strong enough to improve affinity prediction over a well-constructed static residue graph. The thesis contributes a fully audited end-to-end pipeline, a controlled comparison of graph view and feature mode, and a robust negative result that is scientifically useful precisely because it survives repeated held-out evaluation.
\end{abstract}
