\chapter{Discussion and Conclusions}
\label{ch:discussion}

\section{Principal findings}

The final confirmatory matrix leads to one strong positive conclusion and one strong negative conclusion. The positive conclusion is that the full residue graph is consistently more informative than the interface-only graph for the present problem setup. The negative conclusion is that the present molecular-dynamics-derived feature encoding does not provide a robust held-out advantage over the static representation. These conclusions are not based on one favourable split, but on a 52-configuration matrix evaluated over 520 total outer fits on a matched 1437-complex subset.

The strongest overall model was the direct supervised baseline on the full static graph, \texttt{full\_base\_S}, with mean test RMSE 2.3379 and mean test Pearson correlation 0.5064. The best variational model, \texttt{full\_S\_semi\_shared\_static\_z32}, was weaker on both metrics. The best dynamic variational model, \texttt{full\_SD\_semi\_shared\_static\_z8}, came close to the best static variational model, but did not surpass the full static baseline and did not establish a stable matrix-wide advantage for the dynamic feature set. These are not the most optimistic results one might have hoped for at the start of the project, but they are precise, interpretable, and scientifically useful.

\section{Why the full graph outperformed the interface graph}

The superiority of the full graph is one of the most stable results in the thesis. Every matched variational comparison favored the full representation, and the same direction held for the supervised baselines. The most plausible explanation is that the current interface construction discards too much information. In the final retained set, full graphs contained on average 495.9 nodes and 10\,697.2 edges, whereas interface graphs contained on average 127.3 nodes and 2106.3 edges. More importantly, 1395 of the 1437 interface graphs, or 97.1\%, reached the hard cap of 128 nodes. This means that the interface representation was not merely focusing attention on a small local region; in most cases it was being actively truncated at the patch limit.

That truncation matters because affinity is not a purely local property of the closest contact region. Long-range electrostatics, secondary-shell packing, broader chain geometry, and network effects propagated through the graph can all influence a graph-level predictor. The full graph preserves this context, whereas the interface graph imposes an aggressive bottleneck. The interface patch is also defined operationally, in the final experiments, around the closest inter-chain C$\alpha$ pair from the molecular-dynamics-mapped coordinates. This is a sensible local anchor, but it is still a heuristic. If the most predictive residues are not fully captured by that local expansion, the resulting graph will be intrinsically information-limited.

The practical conclusion is therefore not that interface modelling is inherently weak. It is that the specific 128-node interface patch used here is too restrictive to compete with the full graph on the matched PPB-Affinity subset. Any future claim that interface-only graphs should dominate would need either a richer interface definition or evidence from a different and less saturating patching strategy.

\section{Why the present dynamic features did not help}

The dynamic result requires more careful interpretation. It would be too strong to conclude that dynamics are unimportant for protein--protein affinity. What the results support is a narrower statement: the present dynamic representation does not add a robust held-out benefit over the static graph. Several aspects of the implementation help explain why.

First, the trajectory budget is short. Each complex contributes 1.2 ns of saved trajectory, of which 0.2 ns are equilibration frames and 1.0 ns are additional production frames. This is enough to estimate local fluctuations and contact persistence at a coarse level, but it is not long enough to explore major conformational rearrangements for most protein complexes. Second, the trajectory is summarized very aggressively. Hydration is reduced to a per-residue mean and standard deviation of nearby water counts, flexibility is reduced to C$\alpha$ RMSF, torsional state is reduced to mean $\sin/\cos$ values, and edge dynamics are reduced to contact frequency and distance moments on a fixed pre-existing edge set. Temporal order, slow state switching, and changing graph topology are all discarded.

Third, the force features are only approximate proxies. They are based on reconstructed protein-only analysis systems and direct ligand--receptor Coulomb and Lennard--Jones interaction groups, not on exact force decomposition from the full solvated PME simulation context. As a consequence, they are physically interpretable but only at the level of coarse direct inter-chain nonbonded magnitude summaries. Fourth, the simulation metadata are imperfect. Missing residues are not rebuilt, non-protein components are removed before simulation, trajectories include equilibration frames, and pH defaults to 7.0 because pH metadata are absent from the dataset. Experimental temperature is also frequently unavailable and therefore defaults to 298.15 K in many cases. None of these choices invalidates the pipeline, but together they make it easier for the dynamic channels to behave as noisy covariates rather than as stable predictive signals.

Seen in this light, the negative dynamic result is scientifically coherent. The dynamic channels are physically motivated, but coarse; informative, but imperfect; and expensive, but not rich enough to dominate a strong static graph. The most defensible conclusion is therefore that the current dynamic encoding is not yet a better representation than the static one, not that dynamics themselves are irrelevant.

\section{Why the direct supervised baseline was strongest}

The direct supervised baseline outperformed the best variational model by a modest but consistent margin. This outcome is plausible. The supervised baseline devotes all of its capacity to predicting affinity, whereas the GraphVAE must balance reconstruction, latent regularization, and optional supervision. The variational model is therefore solving a harder optimization problem with more ways to spend capacity on objectives that are only indirectly aligned with held-out affinity prediction.

The results suggest that this trade-off is not favourable at the present data scale. On the matched 1437-complex subset, the full static baseline can exploit the graph representation directly and efficiently. The GraphVAE, by contrast, learns useful latent spaces, but its best latent representation still does not outperform the direct predictive model when judged on held-out RMSE and correlation. This does not mean the GraphVAE is without value. It still offers an interpretable latent representation, a principled reconstruction objective, and a platform for future representation-learning experiments. It does mean, however, that for the present thesis question, the simpler predictive model is the stronger empirical answer.

\section{What the train--validation--test behaviour shows}

The overfitting patterns help distinguish genuine representation weakness from a mere lack of model capacity. The unsupervised GraphVAE families exhibit small train--test gaps, but this is largely because they underfit the affinity target even on their training sets. The semi-supervised families reveal a clearer story. In particular, the semi-supervised $SD$ families fit the training data more strongly than the corresponding $S$ families, yet do not improve held-out performance. This is exactly the pattern one would expect when a feature set increases model flexibility more than it increases robust signal.

The most informative comparison is between the full semi-supervised static and full semi-supervised dynamic families. The dynamic family achieves higher mean training correlation, but lower mean test correlation and a substantially larger train--test RMSE gap. The same pattern is even stronger for the interface semi-supervised dynamic family. In other words, the dynamic channels are not merely neutral; they appear to provide additional degrees of freedom that the model can exploit on the training set without converting them into better held-out predictions. This is why the overfitting story is central to interpreting the dynamic results.

\section{Limitations and threats to validity}

Several limitations should be stated explicitly. The first is dataset coverage. Although the curated PPB-Affinity subset contained 2818 complexes, the final confirmatory analysis was restricted to a matched retained subset of 1437 complexes after molecular-dynamics processing, force-feature compatibility checks, and interface-view materialization. The results are therefore not estimates on the entire curated dataset, but on the subset that the full active pipeline could process consistently.

The second limitation is physical fidelity. Missing residues are not rebuilt, heterogens are removed, all simulations use protein-only prepared systems in explicit solvent, and some experimental metadata are missing. The dynamic descriptors are therefore derived from physically meaningful but simplified systems. The third limitation is representational fidelity. The graph topology is static, the interface patch is heavily capped, the trajectories are short, and the dynamic descriptors are summary statistics rather than learned temporal representations. The fourth limitation is evaluation scope. Although the repeated outer resampling design is much stronger than a single split, it is still an internal validation on one dataset family rather than an external benchmark on an entirely different source of complexes and affinity measurements.

These limitations do not invalidate the thesis conclusions. In fact, the main conclusions are carefully phrased so that they remain valid under these limitations. What they do restrict is the scope of interpretation. The thesis provides evidence about a specific and fully described MD-to-graph pipeline. It does not provide the final word on whether dynamics can help protein--protein affinity prediction in general.

\section{Future work}

The most immediate improvement would be to revise the dynamic feature construction rather than to abandon dynamics altogether. Excluding equilibration frames from the feature summaries would be a straightforward correction. Longer production trajectories would make the dynamic summaries less dominated by local relaxation and more sensitive to genuine conformational variability. A richer representation could also retain temporal structure instead of collapsing trajectories to means and standard deviations. Examples include time-windowed summaries, recurrent or temporal graph architectures, or graph construction that changes with the trajectory rather than remaining fixed.

On the graph side, the interface result suggests that the current patching strategy is too restrictive. A larger interface budget, a soft attention mechanism over full graphs, or a hierarchical model that combines interface-local and global context could test whether the weakness lies in the concept of interface focus or in the present hard truncation. On the modelling side, the direct supervised baseline sets a demanding but useful benchmark. Any future representation-learning model should aim either to beat this baseline directly or to offer a secondary benefit, such as transferability or interpretability, that justifies its additional complexity.

\section{Conclusion}

The thesis set out to test whether short explicit-solvent molecular dynamics could improve residue-graph prediction of protein--protein binding affinity on PPB-Affinity-derived complexes. The final answer is precise. For the implemented pipeline, the full residue graph is better than the interface patch, the current dynamic feature set does not improve robust held-out performance over the static graph, and the strongest overall model is a direct supervised baseline on the full static graph. These findings are supported by a repeated outer held-out evaluation rather than by a single favourable split.

This is a useful outcome. It demonstrates that careful benchmarking can overturn optimistic first impressions, that negative results can still be informative when they are rigorously established, and that strong static baselines remain difficult to beat. Just as importantly, it leaves a clear path forward. The thesis does not show that dynamics are irrelevant. It shows that dynamics must be represented more carefully than they were here if they are to improve affinity prediction in a reliable way.
