\chapter{Introduction}
\label{ch:intro}

\section{Problem setting and motivation}

Protein--protein interactions are central to many biological processes and have become increasingly relevant as therapeutic targets \citep{Modell2016PPIs}. In many scientific and therapeutic settings, the key question is not simply whether two proteins bind, but how strongly they bind. The relevant thermodynamic quantity is the binding free energy, or equivalently the dissociation constant $K_D$, which determines how readily a bound complex forms and dissociates under equilibrium conditions. Predicting this quantity from structure is attractive because it is experimentally measurable and directly relevant to biomolecular design \citep{Guo2022PPBAReview,Liu2024PPBAffinity}. It is also difficult. The measured affinity reflects more than a single static interface geometry, while the available structure--affinity datasets for protein--protein complexes remain relatively limited and heterogeneous \citep{Guo2022PPBAReview,Liu2024PPBAffinity}.

The recent PPB-Affinity dataset was created specifically to support machine-learning research on protein--protein binding affinity prediction \citep{Liu2024PPBAffinity}. It provides a scale and level of curation that make systematic benchmarking possible. At the same time, the existence of a large dataset does not remove the underlying modelling challenges. A static crystal or cryo-EM structure captures only one experimentally observed conformation of a complex, whereas binding is a thermodynamic process taking place over an ensemble of solvated molecular states. This motivates a central question for the present thesis: can short molecular-dynamics simulations provide useful additional information beyond a carefully engineered static structural graph?

Graph-based learning is a natural way to pose this question. A protein complex can be represented as a residue-level graph in which residues form nodes and spatially or physicochemically related pairs form edges. Static structural features can then be assigned to nodes and edges, and graph neural networks can be used to learn graph-level embeddings for affinity prediction \citep{Crocioni2024DeepRank2,Xu2019GIN,Hu2020Pretraining,Fey2019PyG}. Such methods are attractive because they can combine local chemical information with higher-order structural context. However, they are still typically built from a single structure or from features extracted from a single frame. Proteins are not rigid objects, and the importance of conformational fluctuations and dynamic ensembles in biomolecular function has long been recognized \citep{HenzlerWildman2007Dynamic}. Hydration, contact persistence, and backbone flexibility may therefore carry predictive information not present in a static graph alone.

The appeal of dynamics comes with a difficulty. Rich trajectory data are expensive to generate, high-dimensional to represent, and easy to summarize badly. It is therefore not enough to ask whether ``dynamics help'' in the abstract. One must ask whether a specific dynamic representation, with a specific simulation budget and a specific graph-learning pipeline, improves held-out predictive performance relative to a strong static baseline. That is the problem addressed here.

\section{Research questions and scope}

This thesis focuses on the final active pipeline implemented in the repository rather than on earlier exploratory branches that were later archived. The workflow begins with a curated broad subset of PPB-Affinity, prepares each complex for explicit-solvent molecular dynamics, constructs residue graphs with DeepRank2, augments these graphs with molecular-dynamics-derived descriptors, and trains either a masked graph variational autoencoder or a direct supervised graph regressor. The evaluation is restricted to a matched subset of complexes for which both the full-complex graph and the interface-only graph can be materialized successfully, allowing a controlled comparison of graph view and feature mode.

Within this scope, the thesis addresses four specific questions. First, is a full residue graph more predictive than a local interface graph for the present dataset and feature set? Second, do molecular-dynamics-derived node and edge summaries improve predictive performance relative to static graph features alone? Third, does a representation-learning approach based on a masked graph variational autoencoder outperform a direct supervised graph regressor when both are given the same graph inputs? Fourth, what can be learned from the train/validation/test behaviour of these models about underfitting, overfitting, and the reliability of conclusions drawn from one-off train/test splits?

These questions are deliberately narrower than the general question of whether molecular dynamics can ever help affinity prediction. The thesis does not attempt to compute rigorous free energies, perform long-timescale conformational sampling, or exhaustively benchmark all possible graph constructions. Instead, it studies a well-defined and fully reproducible pipeline and asks what can be concluded about that pipeline from a careful confirmatory evaluation.

\section{Contribution of the thesis}

The first contribution of the thesis is engineering: a complete end-to-end pipeline from curated structure--affinity data to graph-based prediction. This includes automated preprocessing of protein complexes, explicit-solvent molecular dynamics in OpenMM \citep{Eastman2024OpenMM8}, residue-level graph construction using DeepRank2 \citep{Crocioni2024DeepRank2}, dynamic feature extraction with MDTraj \citep{McGibbon2015MDTraj}, and two downstream model families. The second contribution is methodological: the thesis compares full and interface graph views, static and static-plus-dynamic feature sets, unsupervised and semi-supervised variational training, and direct supervised baselines under a repeated outer resampling design rather than relying on a single split. The third contribution is scientific: the final results provide a robust negative result for the current dynamic encoding, while also clarifying that the negative result arises from the present representation choice rather than from a principled argument that dynamics are irrelevant.

This last point is important. In a project of this kind, a carefully established negative result is often more informative than an optimistic but fragile positive one. A claim that dynamic features help on one train/test split can be overturned by resampling. A claim that they do not help in a specific, audited pipeline is narrower, but more trustworthy. The thesis therefore aims not only to build a predictive model, but also to demonstrate sound scientific reasoning about what the model can and cannot support.

\section{Thesis structure}

Chapter~\ref{ch:litreview} places the thesis in the context of existing work on protein--protein affinity prediction, molecular simulation, structural learning, and variational representation learning. Chapter~\ref{ch:theory} then provides the technical background needed for the rest of the thesis, including molecular dynamics, residue graphs, graph neural networks, and variational autoencoders. Chapter~\ref{chap:methods} describes the implemented pipeline in replication-level detail, including dataset curation, simulation settings, graph construction, model definitions, and the exact confirmatory resampling protocol. Chapter~\ref{ch:results} reports the final 52-configuration evaluation matrix and quantifies the effects of graph view, feature mode, supervision, target policy, and model family. Chapter~\ref{ch:discussion} interprets these findings, discusses their limitations, and states the main conclusions of the thesis. The appendices provide additional technical explanation and a map from the written thesis to the concrete code and data artifacts in the repository.
