\chapter{Literature Review}
\label{ch:litreview}

\section{Protein--protein affinity prediction as a distinct modelling problem}

Predicting protein--protein binding affinity occupies an awkward position between structural bioinformatics, computational chemistry, and machine learning. It is structurally informed, because the arrangement of residues at and around the interface plainly matters. It is also thermodynamic, because the target quantity is the free energy of association rather than a purely geometric score. That dual character makes the problem harder than many standard structure-to-property tasks. A crystal or cryo-EM complex gives one experimentally observed structure, but the measured affinity reflects a balance of enthalpic and entropic effects across an ensemble of bound and unbound states. Solvent, conformational heterogeneity, and local flexibility all matter, yet they are only partially visible in a single deposited structure.

The practical literature reflects this difficulty. In protein--protein learning, a great deal of work has historically focused on interface classification, docking-model quality assessment, hotspot prediction, or mutation-induced changes in binding energy, rather than direct prediction of absolute binding affinity for a complex. The popularity of SKEMPI-type benchmarks illustrates this point: curated mutation datasets made the supervised prediction of $\Delta \Delta G$ more tractable than the direct prediction of absolute $\Delta G$ for arbitrary complexes \citep{Jankauskaite2019SKEMPI}. Absolute affinity prediction has therefore had a thinner benchmark ecosystem, a narrower set of directly comparable models, and less methodological convergence than adjacent tasks.

This context matters for the present thesis. The question is not simply how to squeeze another fraction of a kcal/mol out of a leaderboard. The more basic question is what representation and what modelling compromise are scientifically reasonable when one wants to predict protein--protein affinity from structure while still keeping the pipeline computationally feasible and auditable.

\section{Physics-based routes to binding free energies}

The most physically direct route to binding affinity is to estimate free-energy differences from molecular simulation. In principle, this is attractive because it targets the quantity of interest directly rather than asking a learned model to approximate it indirectly. In practice, however, the available methods span a wide range of cost, approximation, and implementation complexity.

At the more rigorous end are enhanced-sampling and alchemical approaches. Umbrella sampling is a classic example. It introduces biasing potentials along a chosen reaction coordinate so that rare but important configurations are sampled more often, after which the unbiased free-energy profile is reconstructed \citep{Torrie1977Umbrella}. Alchemical binding free-energy methods go further by defining nonphysical thermodynamic pathways between states and integrating free-energy changes along those paths. These methods have become increasingly mature, especially in protein--ligand problems, and they can be highly informative when carefully set up \citep{Cournia2017RBFE}. Their drawbacks are equally well known: they are expensive, sensitive to protocol design, require substantial expertise, and depend strongly on sampling quality and force-field fidelity. For large protein--protein complexes, these difficulties are magnified by system size, interface flexibility, solvent exposure, and the challenge of defining robust collective variables or alchemical pathways.

At the cheaper end are endpoint methods such as MM/PBSA and MM/GBSA. These combine molecular mechanics energies with continuum-solvent approximations and are widely used as approximate binding estimators \citep{Genheden2015MMGBSA}. Their appeal is obvious: they are much cheaper than full free-energy calculations and easy to layer onto existing MD workflows. Their limitations are equally important. They depend on modelling assumptions about solvation, entropy, and state averaging, and their absolute predictive accuracy can vary markedly across systems and protocols \citep{Genheden2015MMGBSA}. In other words, they are often useful comparison points, but they do not remove the underlying difficulty of the affinity problem.

This physics literature influenced the design of the present thesis in two ways. First, it motivates taking molecular dynamics seriously: if conformational fluctuations and solvent behaviour matter physically, then a purely static model is at least open to question. Second, it also motivates restraint. Full rigorous free-energy calculations are not a realistic centrepiece for a thesis whose main goal is to compare learned structural representations at scale. That is why the present work uses MD as a source of descriptors rather than as a direct free-energy engine.

\section{Empirical and structure-based models for protein--protein affinity}

Because rigorous simulation is expensive and endpoint methods remain approximate, a parallel literature has pursued empirical or machine-learned affinity estimation from structure. One representative example is PRODIGY, which predicts protein--protein binding affinity from interfacial contact counts and related structural descriptors \citep{Vangone2015Contacts}. The attraction of such models is that they are fast, interpretable, and easy to apply across large datasets. The drawback is that they compress a complex physical process into a small number of hand-designed descriptors. That can work surprisingly well in some settings, but it also limits the expressive capacity of the model.

The recent PPB-Affinity dataset is an important development in this space because it was assembled specifically to support AI-based protein--protein affinity prediction \citep{Liu2024PPBAffinity}. Its existence signals that the field has moved beyond isolated small benchmarks and toward more systematic evaluation. At the same time, the dataset paper makes clear why the task remains challenging: the available structures are heterogeneous, the complexes span multiple biological regimes, and the number of experimentally labelled cases remains small compared with typical deep-learning scales \citep{Liu2024PPBAffinity}. The problem is therefore large enough to motivate machine learning, but still data-limited enough that representation choice matters acutely.

Recent geometric-learning frameworks such as DeepRank2 show that protein interfaces can be represented effectively with residue-level or atom-level graphs equipped with biophysical node and edge features \citep{Crocioni2024DeepRank2}. That literature is highly relevant even when it is not aimed exclusively at absolute affinity prediction, because it demonstrates a general shift in structural bioinformatics away from rigid handcrafted scoring functions and toward learned geometric representations of macromolecular structure. In this thesis, that shift is taken seriously, but with a narrower question in mind: whether a residue-graph representation augmented with short-trajectory descriptors can outperform an otherwise comparable static representation.

\section{Why the protein--ligand literature is ahead}

A useful way to understand the state of protein--protein affinity prediction is to compare it with the protein--ligand case. Protein--ligand affinity prediction is still difficult, but it benefits from a much longer benchmarking tradition, more standardized datasets, and a clearer methodological pipeline from structure to score. PDBbind is the obvious example of this infrastructure: it provides a large collection of protein--ligand complexes with associated binding measurements and has become a standard reference point for supervised affinity modelling \citep{Wang2004PDBbind}. Once such infrastructure exists, it becomes easier to compare classical scoring functions, 3D convolutional models, graph networks, and MD-enhanced methods on roughly common ground.

This difference in ecosystem partly explains why the protein--ligand literature has moved faster on learned structural models. Pafnucy is a well-known example of a deep learning model that uses a regularized spatial representation to predict protein--ligand binding affinity \citep{Stepniewska2018Pafnucy}. More recently, spatio-temporal learning directly on MD-derived protein--ligand data has also been explored, including models that use trajectory information rather than a single static input \citep{Libouban2025Spatiotemporal}. The protein--ligand case therefore provides a proof of principle that geometric and even trajectory-aware learning can be worthwhile when the data and the representation support it.

Protein--protein complexes are harder in several practical respects. The interfaces are larger, the partners are flexible on both sides, the systems are more heterogeneous, and the data are less standardized. Even the choice of representation is less obvious: a ligand has a natural identity as a small graph or 3D object inside a binding pocket, whereas a protein--protein complex raises immediate questions about whether to represent the whole assembly, only the interface, or some fixed local patch. This is one reason the present thesis compares full-complex and interface-only views directly rather than treating the interface representation as self-evidently correct.

\section{Variational autoencoders in molecular and structural learning}

Variational autoencoders were introduced as a way to learn continuous latent-variable models by combining stochastic encoding, reconstruction, and variational regularization \citep{Kingma2014VAE}. In molecular machine learning, they became attractive not only for generation but also for representation learning: a VAE can force a model to compress high-dimensional structured inputs into a smoother latent space while still preserving information needed for reconstruction or downstream prediction. This idea has been influential in molecular design and molecular graph modelling, for example in continuous latent-space approaches to chemical generation \citep{GomezBombarelli2018MolecularVAE} and in graph-based variational models such as GraphVAE and variational graph autoencoders \citep{Kipf2016VGAE,Simonovsky2018GraphVAE}.

What matters for the present thesis is not molecular generation itself, but the representational role of the VAE. Protein--protein affinity datasets are not large enough to make unrestricted end-to-end supervised learning an obviously optimal choice in every regime. A variational graph model offers a different compromise. It can use reconstruction as an auxiliary task, exploit unlabeled or weakly labeled structure in the data representation, regularize the latent space, and still support downstream supervised evaluation. In other words, the VAE is appealing precisely when one suspects that the input representation contains more structure than the label budget alone can reliably exploit.

At the same time, VAEs are not the dominant standard in direct protein--protein affinity prediction. The literature on affinity prediction is still led more by supervised regression models than by latent-variable approaches. That makes the present thesis somewhat exploratory in how it uses the VAE. The model is not introduced because the literature has already converged on VAEs as the canonical tool for this task. It is introduced because variational representation learning offers a principled way to test whether graph reconstruction and affinity prediction can share a useful latent structure when the labels are limited and the feature space is rich.

\section{Where this thesis fits}

Taken together, the literature suggests a gap rather than a settled recipe. Purely physics-based methods are principled but often too costly or too approximate to serve as the sole practical solution at the scale of modern structural datasets. Purely empirical structure-only scorers are cheap and sometimes effective, but they may compress away information that matters. Protein--ligand work shows that geometric and even trajectory-aware learning can succeed when the data and benchmark culture are strong enough, but the protein--protein case remains less mature. Variational latent models are established tools for molecular representation learning, but they are not yet a standard answer to protein--protein affinity prediction.

The thesis sits in that gap. It does not try to beat rigorous free-energy methods at their own objective, and it does not assume that a single static structural graph must already contain everything worth knowing. Instead, it asks a more focused question: if one starts from a strong residue-graph representation of a protein complex, can short explicit-solvent molecular dynamics be distilled into graph-compatible descriptors that improve held-out affinity prediction? The choice of a graph representation, the inclusion of dynamic summaries, the comparison between full and interface graph views, and the use of a masked GraphVAE alongside a direct supervised baseline all follow naturally from that question. The literature does not dictate that this pipeline should work. It does, however, make the question worth asking.
